%% ----------------------------------------------------------------
%% Thesis.tex -- MAIN FILE (the one that you compile with LaTeX)
%% ----------------------------------------------------------------



% Set up the document
\documentclass[a4paper, 11pt, oneside]{Style/UCLThesis}  % Use the "UCLThesis" style, based on the ECS Thesis style by Steve Gunn

%TC:macro \cmt [ignore]
%TC:macro \note [ignore]
\graphicspath{{Figures/}}  % Location of the graphics files (set up for graphics to be in PDF format)


% Include any extra LaTeX packages required
\usepackage[square, numbers, comma, sort&compress]{natbib}  % Use the "Natbib" style for the references in the Bibliography
\usepackage{verbatim}  % Needed for the "comment" environment to make LaTeX comments
\usepackage{Style/vector}  % Allows "\bvec{}" and "\buvec{}" for "blackboard" style bold vectors in maths
\hypersetup{urlcolor=blue, colorlinks=true}  % Colours hyperlinks in blue, but this can be distracting if there are many links.
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{ltablex}
\usepackage{setspace}
%\usepackage{fullpage}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{color}
%\usepackage{dsfont}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{dirtree}
\usepackage{texshade}
\usepackage{subfigure}
\usepackage{graphicx,picture,calc}
\usepackage{wrapfig}
%\usepackage{lmodern} % Package appears to prevent {verbatim} and {listings} from displaying correctly
\usepackage{lscape}
\usepackage{tikz}
\usepackage{pifont}
\usepackage{fancyvrb}
%\usepackage{textcomp}
%\usepackage{array}

%\listfiles    % useful to check version of packages in the log

%% ----------------------------------------------------------------

\newcommand{\note}[1]{{\color{red} [*** Comment: #1 ***]}}

\begin{document}
\frontmatter	  % Begin Roman style (i, ii, iii, iv...) page numbering

% Set up the Title Page
\title  {Predicting and Characterising Zinc Metal Binding Sites in Proteins}
\authors  {\texorpdfstring
          {Sam M. Ireland}
          {Sam M. Ireland}
          }
\addresses  {\groupname\\\deptname\\\univname}  % Do not change this here, instead these must be set in the "Thesis.cls" file, please look through it instead
\date       {\today}
\subject    {}
\keywords   {}

\maketitle
%% ----------------------------------------------------------------

\setstretch{1.5}  % It is better to have smaller font and larger line spacing than the other way round

% Define the page headers using the FancyHdr package and set up for one-sided printing
\fancyhead{}  % Clears all page headers and footers
\rhead{\thepage}  % Sets the right side header to show the page number
\lhead{}  % Clears the left side page header

\pagestyle{fancy}  % Finally, use the "fancy" page style to implement the FancyHdr headers

%% ----------------------------------------------------------------
% Declaration Page required for the Thesis, your institution may give you a different text to place here
\Declaration{

\addtocontents{toc}{\vspace{1em}}  % Add a gap in the Contents, for aesthetics

I, SAM IRELAND, declare that this thesis titled, `Predicting and Characterising Zinc Metal Binding Sites in Proteins' and the work presented in it are my own. I confirm that:

\begin{itemize}
\item[\tiny{$\blacksquare$}] This work was done wholly or mainly while in candidature for a research degree at this University.

\item[\tiny{$\blacksquare$}] Where any part of this thesis has previously been submitted for a degree or any other qualification at this University or any other institution, this has been clearly stated.

\item[\tiny{$\blacksquare$}] Where I have consulted the published work of others, this is always clearly attributed.

\item[\tiny{$\blacksquare$}] Where I have quoted from the work of others, the source is always given. With the exception of such quotations, this thesis is entirely my own work.

\item[\tiny{$\blacksquare$}] I have acknowledged all main sources of help.

\item[\tiny{$\blacksquare$}] Where the thesis is based on work done by myself jointly with others, I have made clear exactly what was done by others and what I have contributed myself.
\\
\end{itemize}


Signed:\\
\rule[1em]{25em}{0.5pt}  % This prints a line for the signature

Date:\\
\rule[1em]{25em}{0.5pt}  % This prints a line to write the date
}
\clearpage  % Declaration ended, now start a new page

\begin{comment}
%% ----------------------------------------------------------------
% The "Funny Quote Page"
\pagestyle{empty}  % No headers or footers for the following pages

\null\vfill
% Now comes the "Funny Quote", written in italics
\textit{``A doughnut, my reign for a doughnut !!!''}

\begin{flushright}
*** AUTHOR NAME HERE ***
\end{flushright}

\vfill\vfill\vfill\vfill\vfill\vfill\null
\clearpage  % Funny Quote page ended, start a new page
\end{comment}
%% ----------------------------------------------------------------

% The Abstract Page
\chapter{Impact Statement}

This PhD has resulted in the creation of a dedicated database of zinc binding sites (ZincBindDB) and tools for easily accessing it, predictive models for predicting zinc binding sites in proteins (ZincBindPredict), and a novel Python library for analysing protein structures (atomium).

There was previously no such resource for collating zinc binding sites, and so the introduction of one facilitates research into zinc binding sites generally by giving researchers in this field a convenient, centralised repository of the sum of all knowledge of zinc binding sites for which there is structural information - where no such centralised repository existed before. This underpins and supports research into a diverse area of biology. The codebase is also open source and highly extensible, and could be expanded to support other metals as required.

Zinc binding is a property of about 10\% of proteins, and is particularly important in certain enzyme reactions. Knowing whether a protein binds to zinc can offer insights into its function, and knowing precisely where it binds zinc can show the mechanism by which it carries out its intended function, as well as provide suggestions as to how pharmaceutical molecules might disrupt or enhance this function where required for medical interventions. The tools developed here make it easier to computational assess whether a protein binds zinc, and where it does if so - aiding these endeavours.

The newly developed Python library atomium makes studying and analysing protein structures easier for Python programmers, specifically giving them the ability to study higher order assemblies of protein chains called biological assemblies, which was not possible with the existing BioPython library. This makes structural biology research more efficient.


\clearpage  % Abstract ended, start a new page
%% ----------------------------------------------------------------

% The Abstract Page
\addtotoc{Abstract}  % Add the "Abstract" page entry to the Contents
\abstract{
\addtocontents{toc}{\vspace{1em}}  % Add a gap in the Contents, for aesthetics

Zinc is one of the most important biologically active metals. Ten per cent of the human genome is thought to encode a zinc binding protein and its uses encompass catalysis, structural stability, gene expression and immunity. Knowing whether a protein binds to zinc can offer insights into its function, and knowing precisely where it binds zinc can show the mechanism by which it carries out its intended function, as well as provide suggestions as to how pharmaceutical molecules might disrupt or enhance this function where required for medical interventions. At present, there is no specific resource devoted to identifying and presenting all currently known zinc binding sites. This PhD has resulted in the creation of ZincBind --- a database of zinc binding sites (ZincBindDB), predictive models of zinc binding at the family level (ZincBindPredict) and a user-friendly, modern website frontend (ZincBindWeb). Both ZincBindDB and ZincBindPredict are also available as GraphQL APIs. The database of zinc binding sites currently contains 38,141 sites, and is automatically updated every week. The predictive models, trained using the Random Forest Machine Learning algorithm, all achieve an MCC $\ge$ 0.88, recall $\ge$0.93 and precision $\ge$0.91 for the structural models (mean MCC = 0.97), while the sequence models have MCC $\ge$ 0.64, recall $\ge$0.80 and precision $\ge$0.83 (mean MCC = 0.87), outperforming competing, previous predictive models.
}

\clearpage  % Abstract ended, start a new page
%% ----------------------------------------------------------------

\setstretch{1.3}  % Reset the line-spacing to 1.3 for body text (if it has changed)

% The Acknowledgements page, for thanking everyone
\acknowledgements{
\addtocontents{toc}{\vspace{1em}}  % Add a gap in the Contents, for aesthetics

I am enormously grateful to the Wellcome Trust who, in addition to funding the entirety of this PhD and its associated costs, provided considerable assistance and understanding during the COVID pandemic with their prompt, targeted response.

I am greatly indebted to my supervisor, Professor Andrew Martin. Enormously helpful with questions, feedback and guidance at every stage of this project, while still granting me almost complete oversight of the direction and management of the project itself --- I could not have asked for a friendlier, reliable and domain-knowledgable primary supervisor.

Likewise I am grateful to my second supervisor, Professor Stephen Perkins, and my thesis chair, Professor Christine Orengo, whose feedback and contributions at our many meetings were helpful to the eventual success of the project.

Prior to the start of the project I completed three rotations as part of this programme, and the help, guidance, and above all welcoming nature of those three supervisors --- Professor Stephen Perkins (again), Professor Bonnie Wallace, and Professor Francesco Gervasio, to whom I partly owe the relative ease with which I was able to settle into PhD life --- should be noted.

Indeed I would particularly like to single out the Postdoc Dr. Altin Sula, of Professor Bonnie Wallace's lab, whose patience, knowledge, mentorship and generosity of time was invaluable in making that second rotation the considerable success that it was --- particularly given my relative lack of wet lab experience going into it.

More generally I am grateful to my fellow PhD Students and other colleagues at UCL and the Institute of Structural and Molecular Biology for their help, insights and comradeship.

Finally, I am indebted to the countless researchers and technicians who, over the past five decades, have painstakingly populated the Protein Data Bank with structures. Each of the tens of thousands of structures which a computational biologist such as myself considers for a fraction of a second was the work of months of highly skilled labour by teams of people who I will never meet or know, but by whose endeavours this project and the many like it are made possible. Thanks.

}
\clearpage  % End of the Acknowledgements
%% ----------------------------------------------------------------

\pagestyle{fancy}  %The page style headers have been "empty" all this time, now use the "fancy" headers as defined before to bring them back

%% ----------------------------------------------------------------
\lhead{\emph{Contents}}  % Set the left side page header to "Contents"
\tableofcontents  % Write out the Table of Contents

%% ----------------------------------------------------------------
\lhead{\emph{List of Figures}}  % Set the left side page header to "List of Figures"
\listoffigures  % Write out the List of Figures

%% ----------------------------------------------------------------
\lhead{\emph{List of Tables}}  % Set the left side page header to "List of Tables"
\listoftables  % Write out the List of Tables

%% ----------------------------------------------------------------

%% ----------------------------------------------------------------
\begin{comment}
\clearpage  %Start a new page
\lhead{\emph{Symbols}}  % Set the left side page header to "Symbols"
\listofnomenclature{lll}  % Include a list of Symbols (a three column table)
{
% symbol & name & unit \\
$a$ & distance & m \\
$P$ & power & W (Js$^{-1}$) \\
& & \\ % Gap to separate the Roman symbols from the Greek
$\omega$ & angular frequency & rads$^{-1}$ \\
}

%% ----------------------------------------------------------------
% End of the pre-able, contents and lists of things
% Begin the Dedication page

\setstretch{1.5}  % Return the line spacing back to 1.5

\pagestyle{empty}  % Page style needs to be empty for this page
\dedicatory{For/Dedicated to/To my\ldots}

\addtocontents{toc}{\vspace{2em}}  % Add a gap in the Contents, for aesthetics

\end{comment}

%% ----------------------------------------------------------------
\mainmatter	  % Begin normal, numeric (1,2,3...) page numbering
\pagestyle{fancy}  % Return the page headers back to the "fancy" style

% \includeonly{.....}

\include{Chapter_1}
\include{Chapter_2}
\include{Chapter_3}
\include{Chapter_4}
\include{Chapter_5}
%\include{Chapter_6}
%\include{Chapter_7}


%% ----------------------------------------------------------------
% Now begin the Appendices, including them as separate files

\addtocontents{toc}{\vspace{2em}} % Add a gap in the Contents, for aesthetics

\appendix % Cue to tell LaTeX that the following 'chapters' are Appendices

\include{Appendix_A}	% Appendix Title
\include{Appendix_B}	% Appendix Title


\addtocontents{toc}{\vspace{2em}}  % Add a gap in the Contents, for aesthetics
\backmatter

%% ----------------------------------------------------------------
\label{Bibliography}
\lhead{\emph{Bibliography}}  % Change the left side page header to "Bibliography"
%\bibliographystyle{unsrtnat}  % Use the "unsrtnat" BibTeX style for formatting the Bibliography
\bibliographystyle{ieeetr}  % Use the "unsrtnat" BibTeX style for formatting the Bibliography
\bibliography{Bibliography}  % The references (bibliography) information are stored in the file named "Bibliography.bib"

\end{document}  % The End
%% ----------------------------------------------------------------
